%Acknowledgements
\thispagestyle{empty}
\section*{Acknowledgements}
In undertaking this PhD I've had help from a few people within the department at Glasgow and I use this page to gratefully acknowledge their effort. Firstly, my supervisor Martin Hendry for providing me with the opportunity to undertake the project as well as the support he has given along the way. Secondly, I thank Luis Teodoro for helping to understand the theory behind wave-mechanics and for providing numerous suggestions in order to build a robust code. He was a strong proponent of writing a code that preserved the unitary nature of quantum mechanics and hence conserve key physical values.
\newline
\newline
I'd like to give thanks to the following people for the help they provided along the way (in no particular order):
\newline
\newline
Hamish Reid, my office mate, for interesting and thought provoking discussion over the course of the project and also for helping to optimize my final code.
\newline
\newline
Tobia Carozzi for suggesting a method of finding vorticity in the velocity fields and for strongly encouraging me to calculate total momentum for use as a diagnostic of the code. He also helped to create the idea of particles with structure and the accompanying version of wave-mechanics as presented in the final chapter.
\newline
\newline
David Sutherland for help with the interpretation of wave-mechanics in Chapter 3.
\newline
\newline
Hugh Potts for help with multi-dimensional fourier transforms.
\newline
\newline
I owe many thanks to Rebecca Johnston (Cambridge University), a fellow researcher in LSS-wave-mechanics, for useful discussions on the subject matter.